---
title: "SFE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SFE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SFEWorkshop)
```

OK, I have limited space. What are the most important things users should know about SFE? I think for this example, I'll focus on just one sample, and I'm ready to answer questions about multiple samples. Keep this one to 10 minutes or so.

1. Intro to Simple Feature and operations of the sf data frame
2. Intro to SCE and SPE, so the audience get the background for what SFE is extending
3. Intro to SFE, use that schematic. Mention opportunities presented with spatial information unavailable from non-spatial scRNA-seq
4. Intro to the muscle dataset used

1. Reading Space Ranger output directly as SFE
2. colGeometry getters and setters; rowGeometry is less commonly used
3. annotGeometry getters and setters and use cases
4. Spatial graphs, spdep wrappers, Visium graph
5. Geometric operations to find number of myofibers and nuclei per spot, and to subset the SFE object
5. Just briefly mention multiple samples
