---
title: "Voyager"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Voyager}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SFEWorkshop)
```

What are the most important things I want users to know about Voyager? I think this should be 10 minutes or so. Practice at home first.

1. Plotting gene expression and colData in space
2. Plotting what's in 1 plus annotation geometry
3. Moran's I permutation test and plotting. I think I'll lump this with just plain Moran's I and Geary's C in one section.
4. Correlograms and plotting. Maybe not mentioning clustering if I don't have time. Shall I mention checking whether residuals of regression models have spatial autocorrelation? Maybe not now, but it may be interesting when benchmarking regression based SVG methods. 
5. Moran plot, not mentioning clustering if I don't have time.
6. Briefly mention univariate local spatial autocorrelation metrics, multivariate spatial methods, spatial regression, and limitations regarding spatial point processes, raster, and 3D data. Maybe just keep it to the slides.
